Naturalness and Long-lived Beyond the Standard Model Particles

### 2017-07-28 tex/draft.tex
Most of the typesetting

\section{The CMS Detector}\fxnote{Add pictures of everything}
\subsection{Overview}
    \fxnote{Add LHC description}  
    \fxnote{define eta, coordinate system in general}  
    \fxnote{Mention that ATLAS is similar}  
    \fxnote{Maybe mention PDFs. Definitely know about them}  
    \fxnote{cite cms experiment at the cern lhc}
    The Compact Muon Solenoid (CMS) detector is designed to reconstruct a variety of particles from proton-proton collisions at the Large Hadron Collider (LHC). To meet this goal, CMS combines information from several subdetectors nested radially in and around a \SI{4}{\tesla} superconducting solenoid. All together, CMS is \SI{21.6}{\m} long, \SI{14.6}{\m} across, and weighs approximately \SI{12500}{t}. The subdetectors that make up CMS are described below.

\subsection{Tracker}
    \fxnote{cite CMS TDR, jinst 12 c02033}
    \fxnote{mention that analyses below used pre-upgrade tracker}
    \fxnote{mention stereo modules and be able to explain stereo modules}
    In the region closest to the interaction point, CMS uses a fast, high-granularity silicon tracker to reconstruct tracks and pinpoint primary and secondary vertices.\fxnote{out-of-fielder may not know about 'track' or 'vertex'} The tracker is subdivided into two regions: at $\mathrm{r}<\SI{20}{cm}$, the high particle flux \fxnote{specify} requires small pixels to keep occupancy near \SI{0.1}{\percent}. Moving farther from the IP decreases particle flux, which allows the use of larger silicon strips.

@@

\section{LHC Searches for Long-lived Particles}
    \fxnote{Include diagrams for signal models} 
    \fxnote{mention which data was used for each search} 
    \fxnote{Steal limit plots and any cool figures from papers} 
\subsection{Overview}
    LLPs leave distinct signatures that differentiate LLP signals from SM backgrounds but also add complexity to LLP analyses. CMS and ATLAS were generally designed to detect and reconstruct stable particles from prompt decays, so LLP analyses often need to stray from the traditional analysis path in terms of object reconstruction \fxnote{really reconstruction? Maybe object definition?}, background estimation, and which data tier to use \fxnote{data tier is too jargony}. This paper will focus on four complementary LLP signatures: displaced vertices, disappearing tracks, slow-moving particles, and particles that stop completely before decaying.

    Displaced vertices and disappearing tracks both look for particles that are unstable on detector timescales \fxnote{quantify?}. Displaced vertices searches are sensitive to LLPs that decay to visible particles in the inner volume \fxnote{specify} of the detector and leave tracks that can be traced back to a vertex that is sufficiently displaced from the nominal interaction point \fxnote{wordy}. Disappearing track searches are sensitive to LLPs that decay to particles that are either neutral or too soft to be reconstructed. Together, these two types of analyses cover \fxnote{something something} parameter space of LLPs that decay on detector timescales.

    Slow and stopped particle searches are both sensitive to LLPs that are detector-stable but far more massive than any stable SM particle. Searches for low-velocity particles use ionization energy loss per unit distance (dE/dX) and time-of-flight (TOF) measurements to look for particles that are moving more slowly than the expected SM particles \fxnote{redundant}, which normally \fxnote{quantify} have beta\fxnote{define} $>$ 0.9. Stopped particle searches take this idea one step further and look for particles that actually come to rest inside the detector before decaying, a phenomenon that could be possible with long lifetimes and small initial kinetic energy. These analyses look for calorimeter energy deposits or hits in the muon system that are out of time with proton bunch crossings that could be due to the decay of stopped particles.

@@

\subsection{Displaced Vertices}
    \fxnote{cite CONF-2017-026}
    LLPs that decay before reaching the inner tracker but still live longer than the vast majority of SM particles can be identified with a displaced vertex (DV). That is, a reconstructed vertex far from the nominal interaction point. ATLAS performs a search for such particles by looking for at least one DV along with large MET and large track multiplicity. The search is sensitive to LLPs with lifetimes in the 20 ps to 10 ns range.

@@

    Jets and MET are reconstructed as in standard analyses, but special care must be taken when looking for displaced tracks. The standard ATLAS tracking algorithm constrains the transverse and longitudinal impact parameters of potential tracks, so the authors employ an additional algorithm, referred to as large-radius tracking (LRT). LRT is run after the standard tracking algorithm and uses only hits that were not used in standard track reconstruction. By relaxing d0, z0, and number of hits requirements, LRT is able to reconstruct tracks that point back to DVs.

@@

    After estimating 0.02 +- 0.02 background events, exactly 0 are observed. The result is then interpreted as an upper limit on the gluino mass and pair-production cross section as a function of gluino lifetime. For an LSP neutralino mass of \SI{100}{\giga\electronvolt}, the gluino mass limit is approximately \SI{2000}{\giga\electronvolt} for lifetimes in the 20 ps to 10 ns range. \fxnote{Include plots}

\subsection{Disappearing Tracks}
    \fxnote{cite CONF-2017-017}
    \fxnote{nice disappearing track cartoon}
    \fxnote{nice cartoon feynman diagrams of signal processes}
    Certain models with nearly mass-degenerate LSP and NLSP have the potential to present a remarkable signal: charged particles that seem to disappear in the tracker. Specifically, models with winos as the lightest gauginos, such as many AMSB models, will have a chargino NLSP with an observable lifetime (0.2 ns is a typical estimate) that eventaully decays to the neutralino LSP and, commonly, a pion. The nearly-degenerate masses have two important phenomenological roles: (1) increase the chargino lifetime, and (2) ensure low-momentum decay products. The low-momentum pion is then unlikely to be constructed, which results in a disappearing track.

    Both ATLAS and CMS looked for disappearing tracks during Run I, but the tracker improvements installed in ATLAS before 2015 and CMS before 2017 should increase sensitivity to decays at shorter radii \fxnote{I haven't mentioned previous iterations when talking about other analyses}. The ALTAS search presented here is sensitive to LLP lifetimes of approximated 10 ps to 10 ns \fxnote{make sure I'm talking about lifetime sensitivity in the same way across analyses}.

@@

    After estimating the background by extrapolating data from the cosmic-enriched region to the signal region and determining systematic uncertainties related to the MC simulation, trigger acceptance, and luminosity, the authors calculated the expected background for LLP lifetimes ranging from 100 ns to 1 us. In all cases, the expected background was less then one event. Exactly zero events are observed in every lifetime scenario. These results are then interpreted as upper limits on the signal production cross section\fxnote{and mass limits}\fxnote{include limit plots or table}, which are the first limits for stopped particles that decay to muons at the LHC.

@@

    HSCPs will often be misidentified or unobserved by standard reconstruction algorithms and analysis criteria, which tend to assume beta$\approx$1 and $|q|$ = e. Inverting these assumptions provides two handles for a dedicated HSCP search in the context of three benchmark models: (1) HSCPs that form R-hadrons that either change the sign of their electric charge or become neutral before reaching the MS, (2) lepton-like HSCPs (quasi-stable sleptons) that are produced either through decays of squarks or gluinos or directly if squarks and gluinos are kinematically inaccessible, and (3) long-lived leptonlike fermions with arbitrary electric charge. This discussion will focus on the first two models, as they bear more directly on naturalness and SUSY.

    \fxnote{I could talk about how low-beta tracks will show up as zig-zags in DTs b/c reconstruction assumes beta$\approx$1. Zig-zag offset is used to estimate beta. Not sure where to fit it in.} \fxnote{Worth mentioning tracker-only vs tracker+TOF}

@@

    The authors estimate the signal-region background using an ABCD method in which events are binned in a 2D plane with uncorrelated axes and the ratio of events in two control regions is used to extrapolate into the signal region from a third control region \fxnote{Do I need to explain this? If I do, I should be more clear}. Systematic uncertainties are introduced in background estimation, signal acceptance, and integrated luminosity.

@@

    When building a natural SUSY model, certain constraints are nearly unavoidable\fxnote{Dangerous. What if Linda knows how to avoid them?}. Specifically, the higgsino, which affects mH at tree level; the stop, which affects mH at one loop; and the gluino, which affets mStop at one loop and therefore mH at two, must all be heavy enough to pull mH up to \SI{125}{\giga\electronvolt}\fxnote{Do I really understand this?} but light enough to avoid excessive fine tuning. Furthermore, the stop and gluino masses are correlated -\fxnote{m-dash?} a large difference between the two would be another source of fine tuning.

    Natural SUSY scenarios, then, are most likely to be constrained by limits on stop and gluino masses due to the strong interaction of these two objects as opposed to the colorless Higgsino\fxnote{Dear lord, I don't understand why I can't figure out how to word this sentence}. Demanding naturalness also constrains the messenger scale of SUSY breaking\fxnote{Make sure I understand this. Messenger scale != breaking scale?}. Corrections to the Higgs mass from stops and gluinos depend of the messenger scale\fxnote{reference eqs}, so natural SUSY generally prefers a lower scale.\fxnote{Either explain more thoroughly or drop completely. What are the consequences of a lower messenger scale?}

    Indeed, limits on mStop and mGluino from conventional SUSY searches wipe out much of the natural parameter space\fxnote{quantify? That would require getting into how one quantifies fine tuning} for many models with decoupled first and second geration fermions, baryonic RPV, or compressed spectra that minimize missing energy while the natural MSSM is completely excluded.\fxnote{Good spot for sweet Cornering plots}\fxnote{Use ATLAS SUSY summary plot too. CMS's isn't updated}

>>>>

\section{The CMS Detector}\fxnote{Add pictures of everything}
\subsection{Overview}
    \fxnote{define eta, coordinate system in general}  
    \fxnote{Mention that ATLAS is similar, says Andrew}  
    \fxnote{Maybe mention PDFs. Definitely know about them, says Andrew}  
    \fxnote{cite cms experiment at the cern lhc}
    The Compact Muon Solenoid (CMS) detector is designed to reconstruct a variety of particles from collisions at the Large Hadron Collider (LHC), which is a circular proton-proton collider that ran with center-of-mass energy \SI{7}{\tera\electronvolt} in 2010-11, \SI{8}{\tera\electronvolt} in 2012, and \SI{13}{\tera\electronvolt} from 2015 to the present. To meet this goal, CMS combines information from several subdetectors nested radially in and around a \SI{4}{\tesla} superconducting solenoid. All together, CMS is \SI{21.6}{\m} long, \SI{14.6}{\m} across, and weighs approximately \SI{12500}{t}. The subdetectors that make up CMS are described below.

\subsection{Tracker}
    \fxnote{cite CMS TDR, jinst 12 c02033}
    \fxnote{mention that analyses below used pre-upgrade tracker}
    In the region closest to the interaction point, CMS uses a fast, high-granularity silicon tracker to reconstruct tracks and pinpoint primary and secondary vertices. The tracker is subdivided into two regions: at $\mathrm{r}<\SI{20}{cm}$, the high particle flux \fxnote{specify} requires small pixels to keep occupancy near \SI{0.1}{\percent}. Moving farther from the IP decreases particle flux, which allows the use of larger silicon strips.

@@

\section{LHC Searches for Long-lived Particles}
    \fxnote{Include diagrams for signal models} 
    \fxnote{mention which data was used for each search} 
    \fxnote{Steal limit plots and any cool figures from papers} 
\subsection{Overview}
    LLPs leave distinct signatures that differentiate signal from SM background but also add complexity to LLP analyses. CMS and ATLAS were designed to reconstruct stable particles from prompt decays, so LLP searches are often forced to stray from the traditional analysis path in terms of object reconstruction\fxnote{really reconstruction? Maybe object definition?} and background estimation. This report focuses on four complementary LLP signatures: displaced vertices, disappearing tracks, slow-moving particles, and particles that stop completely before decaying.

    Displaced vertices and disappearing tracks both look for\fxnote{reword. the vertices and tracks don't look for anything} particles that are unstable on detector timescales. Displaced vertex searches are sensitive to LLPs that decay to visible particles in the inner volume \fxnote{specify} of the detector and leave tracks that can be traced back to a vertex that is sufficiently displaced from the nominal interaction point \fxnote{wordy}. Disappearing track searches are sensitive to LLPs that decay to particles that are either neutral or too soft to be reconstructed. Together, these two types of analyses cover \fxnote{something something} parameter space of LLPs that decay on detector timescales.

    Slow and stopped particle searches are both sensitive to LLPs that are detector-stable but far more massive than any stable SM particle. Searches for low-velocity particles use ionization energy loss per unit distance (dE/dX) and time-of-flight (TOF) measurements to look for particles that are moving more slowly than the expected SM particles \fxnote{redundant}, which normally \fxnote{quantify} have $\beta=\frac{v}{c} > \num{0.9}$. Stopped particle searches take this idea one step further and look for particles that actually come to rest inside the detector before decaying, a phenomenon that could be possible with long lifetimes and small initial kinetic energy. These analyses look for calorimeter energy deposits or hits in the muon system that are out of time with proton bunch crossings that could be due to the decay of stopped particles.

@@

\subsection{Displaced Vertices}
    \fxnote{cite CONF-2017-026}
    LLPs that decay before reaching the inner tracker but still live longer than the vast majority of SM particles can be identified with a displaced vertex (DV). That is, a reconstructed vertex far from the nominal interaction point. ATLAS performs a search for such particles by looking for at least one DV along with large MET and large track multiplicity. The search is sensitive to LLPs with lifetimes in the \SI{20}{\pico\s} to \SI{10}{\nano\s} range.

@@

    Jets and MET are reconstructed as in standard analyses, but special care must be taken when looking for displaced tracks. The standard ATLAS tracking algorithm constrains the transverse and longitudinal impact parameters of potential tracks, so the authors employ an additional algorithm, referred to as large-radius tracking (LRT). LRT is run after the standard tracking algorithm and uses only hits that were not used in standard track reconstruction. By relaxing d0, z0,\fxnote{define} and number of hits requirements, LRT is able to reconstruct tracks that point back to DVs.

@@

    After estimating $\num{0.02} \pm \num{0.02}$ background events, exactly zero are observed. The result is then interpreted as an upper limit on the gluino mass and pair-production cross section as a function of gluino lifetime. For an LSP neutralino mass of \SI{100}{\giga\electronvolt}, the gluino mass limit is approximately \SI{2000}{\giga\electronvolt} for lifetimes in the \SI{20}{\pico\s} to \SI{10}{\nano\s} range. \fxnote{Include plots}

\subsection{Disappearing Tracks}
    \fxnote{cite CONF-2017-017}
    \fxnote{nice disappearing track cartoon}
    \fxnote{nice cartoon feynman diagrams of signal processes}
    Certain models with nearly mass-degenerate LSP and NLSP have the potential to present a remarkable signal: charged particles that seem to disappear in the tracker. Specifically, models with winos as the lightest gauginos, such as many AMSB models, will have a chargino NLSP with an observable lifetime (\SI{0.2}{\nano\s} is a typical estimate) that eventaully decays to the neutralino LSP and, commonly, a pion. The nearly-degenerate masses have two important phenomenological roles: (1) increase the chargino lifetime, and (2) ensure low-momentum decay products. The low-momentum pion is then unlikely to be constructed, which results in a disappearing track.

    Both ATLAS and CMS looked for disappearing tracks during Run I, but the tracker improvements installed in ATLAS before 2015 datataking and CMS before 2017 datataking should increase sensitivity to decays at shorter radii \fxnote{I haven't mentioned previous iterations when talking about other analyses}. The ALTAS search presented here is sensitive to LLP lifetimes of approximated \SI{10}{\pico\s} to \SI{10}{\nano\s} \fxnote{make sure I'm talking about lifetime sensitivity in the same way across analyses}.

@@

    After estimating the background by extrapolating data from the cosmic-enriched region to the signal region and determining systematic uncertainties related to the MC simulation, trigger acceptance, and luminosity, the authors calculated the expected background for LLP lifetimes ranging from \SI{100}{\nano\s} to \SI{1}{\micro\s}. In all cases, the expected background was less then one event. Exactly zero events are observed in every lifetime scenario. These results are then interpreted as upper limits on the signal production cross section\fxnote{and mass limits}\fxnote{include limit plots or table}, which are the first limits for stopped particles that decay to muons at the LHC.

@@

    HSCPs will often be misidentified or unobserved by standard reconstruction algorithms and analysis criteria, which tend to assume $\beta \approx \num{1}$ and $\lvert q \rvert = e$. Inverting these assumptions provides two handles for a dedicated HSCP search in the context of three benchmark models: (1) HSCPs that form R-hadrons that either change the sign of their electric charge or become neutral before reaching the MS, (2) lepton-like HSCPs (quasi-stable sleptons) that are produced either through decays of squarks or gluinos or directly if squarks and gluinos are kinematically inaccessible, and (3) long-lived leptonlike fermions with arbitrary electric charge. This discussion will focus on the first two models, as they bear more directly on naturalness and SUSY.

    \fxnote{I could talk about how low-beta tracks will show up as zig-zags in DTs b/c reconstruction assumes $\beta \approx \num{1}$. Zig-zag offset is used to estimate beta. Not sure where to fit it in.} \fxnote{Worth mentioning tracker-only vs tracker+TOF}

@@

    The authors estimate the signal-region background using an ABCD method in which events are binned in a 2-D plane with uncorrelated axes and the ratio of events in two control regions is used to extrapolate into the signal region from a third control region \fxnote{Do I need to explain this? If I do, I should be more clear}. Systematic uncertainties are introduced in background estimation, signal acceptance, and integrated luminosity.

@@

    When building a natural SUSY model, certain constraints are nearly unavoidable\fxnote{Dangerous. What if Linda knows how to avoid them?}. Specifically, the Higgsino\fxnote{careful, there could be multiple Higgsinos. Do I mean 'lightest Higgsino'?}, which affects mH at tree level; the stop\fxnote{mention that stop is more important than other sfermions b/c large yukawa coupling? I'll have to clarify that somewhere}, which affects mH at one loop; and the gluino, which affets mStop at one loop and therefore mH at two, must all be light enough to avoid excessive fine tuning.\fxnote{diagrams?} Furthermore, the stop and gluino masses are correlated --- a large difference between the two would itself be unnatural.

    Natural SUSY scenarios, then, are most likely to be constrained by limits on the stop, gluino, and Higgsino masses. The Higgsino, however, does not carry color charge and is far less likely to be produced at the LHC,\fxnote{things get wordy and repetative after this} which means that stop and gluino limits will be the most important experimental constraints on natural SUSY. Demanding naturalness also constrains the messenger scale of SUSY breaking\fxnote{Make sure I understand this. Messenger scale != breaking scale?}. Corrections to the Higgs mass from stops and gluinos depend of the messenger scale\fxnote{reference eqs}, so natural SUSY generally prefers a lower scale.\fxnote{Either explain more thoroughly or drop completely. What are the consequences of a lower messenger scale?}

    Indeed, limits on mStop and mGluino from conventional SUSY searches wipe out much of the natural parameter space from prompt decays\fxnote{quantify? That would require getting into how one quantifies fine tuning} for many models with decoupled first and second geration fermions, baryonic RPV, or compressed spectra that minimize missing energy while the natural MSSM is completely excluded.\fxnote{Good spot for sweet Cornering plots}\fxnote{Use ATLAS SUSY summary plot too. CMS's isn't updated}
